Nested sampling with multiple scalars

### 2015-09-29 Paper/ms.tex
A bit of writing

Nested Sampling \citep{skilling} is an effective and popular
Monte Carlo algorithm for Bayesian computation and
statistical mechanics
\citep{2009arXiv0906.3544P, 2014PhRvE..89b2302P, 2015arXiv150303404B}.
In a Bayesian inference problem with unknown parameters denoted collectively
by a vector $\xx$, the
posterior distribution for the parameters given data $\data$ (and prior
information $I$) is:
\begin{eqnarray}
p(\xx | \data, I) &=&
\frac{p(\xx | I)p(\data | \xx, I)}{p(\data | I)}\\
&=& \frac{\pi(\xx)L(\xx)}{Z}
\end{eqnarray}
where $\pi(\xx)$ is the prior distribution, $L(\xx)$ is the likelihood
function, and $Z$ is the normalising constant, known as the
``marginal likelihood'' or the ``evidence'':
\begin{eqnarray}
Z &=& \int \pi(\xx) L(\xx) \, \dx.
\end{eqnarray}

Nested Sampling begins by drawing particles from the
prior $\pi(\xx)$ and successively imposing constraints on the value of
the likelihood $L(\xx)$ that compress the prior mass by a
factor that is approximately known.
This enables the calculation of the marginal likelihood
and properties of the posterior, or any other distribution that is intermediate
between the prior and the posterior. For example, we might be interested in
a ``power posterior'' where the likelihood is raised to a power $\beta$:
\begin{eqnarray}
p(\xx; \beta) &=& \frac{\pi(\xx)L(\xx)^\beta}{Z(\beta)}
\end{eqnarray}
The normalisation and posterior samples from this distribution can be obtained
from the original Nested Sampling run. An example application, in the case of
a Bayesian model with a ``gaussian noise'' assumption in the likelihood,
computing $p(\xx; \beta)$ for $\beta \neq 1$ allows us to explore what the
posterior distribution would have been if the noise variance had been greater.

Relative to alternative approaches such as annealing, Nested Sampling continues
to work when the problem contains a {\it phase transition}.

>>>>

Nested Sampling \citep{skilling} is an effective and popular
Monte Carlo algorithm for Bayesian computation and
statistical mechanics
\citep{2009arXiv0906.3544P, 2014PhRvE..89b2302P, 2015arXiv150303404B}.
In a Bayesian inference problem with unknown parameters denoted collectively
by a vector $\xx$, the
posterior distribution for the parameters given data $\data$ (and prior
information $I$) is:
\begin{eqnarray}
p(\xx | \data, I) &=&
\frac{p(\xx | I)p(\data | \xx, I)}{p(\data | I)}\\
&=& \frac{\pi(\xx)L(\xx)}{Z}
\end{eqnarray}
where $\pi(\xx)$ is the prior distribution, $L(\xx)$ is the likelihood
function, and $Z$ is the normalising constant, known as the
``marginal likelihood'' or the ``evidence'':
\begin{eqnarray}
Z &=& \int \pi(\xx) L(\xx) \, \dx.
\end{eqnarray}
It begins by drawing particles from the
prior $\pi(\xx)$ and successively imposing constraints on the value of
the likelihood $L(\xx)$ that compress the prior mass by a
factor that is approximately known.
This enables the calculation of the marginal likelihood $Z$
and properties of the posterior $\pi(\xx)L(\xx)/Z$.

NS also allows us to calculate the properties of any other distribution that
is in some sense intermediate
between the prior and the posterior. For example, we might be interested in
a ``power posterior'' where the likelihood is raised to a power $\beta$:
\begin{eqnarray}
p(\xx; \beta) &=& \frac{\pi(\xx)L(\xx)^\beta}{Z(\beta)}\label{eqn:power_posterior}
\end{eqnarray}
The normalisation and posterior samples from this distribution can be obtained
from the original Nested Sampling run. An example application, in the case of
a Bayesian model with a ``gaussian noise'' assumption in the likelihood,
computing $p(\xx; \beta)$ for $\beta \neq 1$ allows us to explore what the
posterior distribution would have been if the noise variance had been greater.
Relative to alternative approaches such as annealing, Nested Sampling continues
to work when the problem contains a {\it phase transition}.

In statistical mechanics, Equation~\ref{eqn:power_posterior} defines the
family of {\it canonical distributions}, usually written as:
\begin{eqnarray}
p(\xx; \beta) &=& \frac{\pi(\xx)\exp[-\beta E(\xx)}{Z(\beta)}
\end{eqnarray}
where $E(\xx)$ is the energy function. In this context we usually want to
compute $Z(\beta)$ as a function of $\beta$, which is called the
{\it partition function}.
